---
title: "PS270 Final Project"
author: "Isabelle Seiler"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project Proposal 

My research question is how did race related voting impact the election of Barack Obama versus John McCain in 2008? For my final project, I will use the American National Election Studies (ANES) 2008 Time Series Study. This data will help me unpack how racial identity correlates to presidential vote choice. My hypothesis is that Black voters were more likely to elect Barack Obama in the election, while White voters were more likely to vote for John McCain. This hypothesis is plausible because voters are more likely to shape political preference based on their personal racial identity. Additionally, since Obama was the first Black candidate representing a major-party nominee which could have increased Black voters support for Barack Obama.

My main independent variable is respondent race. In the ANES 2008 study, this operates as a categorical variable that identifies respondents with different racial group (White, Black, Hispanic, Asian, etc). On the other hand, the dependent variable is presidential vote choice in 2008 between the two candidates (Barack Obama and John McCain). Within my code, this variable will represent 1 which is a vote for Obama while 0 represents 0 which is a vote for McCain and excludes respondents who did not vote for one of the two major-party candidates.

The observed pattern in the data would provide support for my hypothesis if Black respondents were more likely to support/vote for Obama. However, the hypothesis is not supported if race does not significantly predict vote choice after I control the specific party identification.





```{r}
install.packages("haven")

library(tidyverse)
library(haven)

anes2008 <- read_dta("data/anes_2008.dta")

glimpse(anes2008)

anes2008_small <- anes2008 |>
  select(V085036d, V083310x, V083097) |>
  filter(V085036d %in% c(1,2)) |>
  filter(V083310x %in% c(1,2)) |>
  mutate(
    vote_obama = if_else(V085036d == 1, 1, 0),
    black = if_else(V083310x == 2, 1, 0)
  )

head(anes2008_small)
```


```{r}
library(tidyverse)

anes2008 <- read_csv(file.choose())

anes2008_small <- anes2008 |>
  select(V085036d, V083310x, V083097) |>
  filter(V085036d %in% c(1,2)) |>
  filter(V083310x %in% c(1,2)) |>
  mutate(
    vote_obama = if_else(V085036d == 1, 1, 0),
    black = if_else(V083310x == 2, 1, 0)
  )

head(anes2008_small)
```


